\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3.2\baselineskip}

### 2017-03-27 paper/ms.tex
A few minor edits, commas

These special cases demonstrate that the stochastically-driven simple harmonic
oscillator provides a physically motivated model that is flexible enough to
describe a wide range of stellar variations.
Low $Q \approx 1$ can capture granulation noise and high $Q \gg 1$ is a good
model for asteroseismic oscillations.
In practice, we take a sum over oscillators with different values of $Q$,
$S_0$, and $\omega_0$ to give a sufficient accounting of the power spectrum
stellar time series.
Since this kernel is exactly described by the exponential kernel, the
likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
measurements in $\mathcal{O}(N)$ operations using the \celeriteterm\ method
described in the previous section.

@@

As discussed previously, the \celerite\ method is based upon a method
developed by Rybicki \& Press twenty years ago \citep{Rybicki:1995}.
In the case where the kernel function has a single term of the form given in
\eq{kernel-simple}, it is interesting to compare the performance of \celerite\
and the Rybicki \& Press method.
Their method does not require solving the extended matrix because the matrix
inverse can be calculated analytically but the method is made slightly more
complicated when white noise is included.
We implement their full method and find empirically that it is more
computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$
but we note that the requirement of a single, real, exponential kernel is
extremely restrictive and this model cannot describe quasi-periodic
variability.

@@

\citet{Carter:2009} improved the scaling of Toeplitz methods by introducing a
wavelet-based method for computing a GP likelihood with $\mathcal{O}(N)$
scaling.
This method has been widely applied in the context of exoplanet transit
characterization but it requires evenly spaced observations and the power
spectrum of the process must have the form $S(\omega)\propto \omega^{-1}$ to
gain the computational advantage.
This wavelet method has been demonstrated to improve parameter estimation for
transiting exoplanets \citep{Carter:2009} but these strict requirements make
this method applicable for only a limited set of use cases.

@@

Many other approximate methods for scaling GP inference exist \citep[see, for
example,][and references therein]{Wilson:2015a} and we make no attempt to make
our discussion exhaustive.
The key takeaway here is that \celerite\ provides an \emph{exact} method for
GP inference for a specific but flexible class of one-dimensional kernel
functions.
Furthermore, since \celeriteterm\ models can be interpreted as a mixture of
stochastically-driven, damped simple harmonic oscillators, they are a
physically motivated choice of covariance function in many applications.

@@

Gaussian Process models have been fruitfully applied to many problems in
astronomical data analysis but the fact that the computational cost scales as
the cube of the number of data points has limited their use to relatively
small datasets.
With the linear scaling of \celerite\, we envision the application of Gaussian
processes may be expanded to much larger datasets.
Despite the restrictive form of the \celeriteterm\ kernel, with a sufficient
number of components it is flexible enough to describe a wide range of
astrophysical variability.
In fact, the relation of the \celeriteterm\ kernel to the damped, stochastically-driven harmonic
oscillator matches simple models of astrophysical variability, and makes the
parameterization interpretable in terms of resonant frequency, amplitude, and
quality factor.
% Siva: Removed major, since $J$ is most often small as stated in the second half of the sentence
A drawback of this method is its quadratic scaling with the number of
terms $J$ but, in many cases, small values of $J$ are sufficient.

@@

The \celeriteterm\ formalism can also be used for power spectrum estimation and
quantification of its uncertainties.
In principle, a large number of \celeriteterm\ terms could be used to perform
non-parametric probabilistic inference of the power spectrum despite
unevenly-spaced data with heteroscedastic noise \citep[for
example,][]{Wilson:2013, Kelly:2014}.
This type of analysis will be limited by the quadratic scaling of \celerite\
with the number of terms $J$ but this limits existing methods as well
\citep[CARMA models,][]{Kelly:2014}.
In general, we encourage the use of physically motivated models for parameter
estimation instead of qualitative modeling of the power spectrum itself.

>>>>

These special cases demonstrate that the stochastically-driven simple harmonic
oscillator provides a physically motivated model that is flexible enough to
describe a wide range of stellar variations.
Low $Q \approx 1$ can capture granulation noise and high $Q \gg 1$ is a good
model for asteroseismic oscillations.
In practice, we take a sum over oscillators with different values of $Q$,
$S_0$, and $\omega_0$ to give a sufficient accounting of the power spectrum
of stellar time series.
Since this kernel is exactly described by the exponential kernel, the
likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
measurements in $\mathcal{O}(N)$ operations using the \celeriteterm\ method
described in the previous section.

@@

As discussed previously, the \celerite\ method is based upon a method
developed by Rybicki \& Press twenty years ago \citep{Rybicki:1995}.
In the case where the kernel function has a single term of the form given in
\eq{kernel-simple}, it is interesting to compare the performance of \celerite\
and the Rybicki \& Press method.
Their method does not require solving the extended matrix because the matrix
inverse can be calculated analytically, but the method is made slightly more
complicated when white noise is included.
We implement their full method and find empirically that it is more
computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$,
but we note that the requirement of a single, real, exponential kernel is
extremely restrictive and this model cannot describe quasi-periodic
variability.

@@

\citet{Carter:2009} improved the scaling of Toeplitz methods by introducing a
wavelet-based method for computing a GP likelihood with $\mathcal{O}(N)$
scaling.
This method has been widely applied in the context of exoplanet transit
characterization, but it requires evenly spaced observations and the power
spectrum of the process must have the form $S(\omega)\propto \omega^{-1}$ to
gain the computational advantage.
This wavelet method has been demonstrated to improve parameter estimation for
transiting exoplanets \citep{Carter:2009}, but these strict requirements make
this method applicable for only a limited set of use cases.

@@

Many other approximate methods for scaling GP inference exist \citep[see, for
example,][and references therein]{Wilson:2015a} and we make no attempt to make
our discussion exhaustive.
The key takeaway here is that \celerite\ provides an \emph{exact} method for
GP inference for a specific, but flexible, class of one-dimensional kernel
functions.
Furthermore, since \celeriteterm\ models can be interpreted as a mixture of
stochastically-driven, damped simple harmonic oscillators, they are a
physically motivated choice of covariance function in many applications.

@@

Gaussian Process models have been fruitfully applied to many problems in
astronomical data analysis, but the fact that the computational cost scales as
the cube of the number of data points has limited their use to relatively
small datasets.
With the linear scaling of \celerite\, we envision the application of Gaussian
processes may be expanded to much larger datasets.
Despite the restrictive form of the \celeriteterm\ kernel, with a sufficient
number of components it is flexible enough to describe a wide range of
astrophysical variability.
In fact, the relation of the \celeriteterm\ kernel to the damped, stochastically-driven harmonic
oscillator matches simple models of astrophysical variability, and makes the
parameterization interpretable in terms of resonant frequency, amplitude, and
quality factor.
% Siva: Removed major, since $J$ is most often small as stated in the second half of the sentence
A drawback of this method is its quadratic scaling with the number of
terms $J$, but, in many cases, small values of $J$ are sufficient.

@@

The \celeriteterm\ formalism can also be used for power spectrum estimation and
quantification of its uncertainties.
In principle, a large number of \celeriteterm\ terms could be used to perform
non-parametric probabilistic inference of the power spectrum despite
unevenly-spaced data with heteroscedastic noise \citep[for
example,][]{Wilson:2013, Kelly:2014}.
This type of analysis will be limited by the quadratic scaling of \celerite\
with the number of terms $J$, but this limits existing methods as well
\citep[CARMA models,][]{Kelly:2014}.
In general, we encourage the use of physically motivated models for parameter
estimation instead of qualitative modeling of the power spectrum itself.
